I Compensate, therefore I Am \\ (accurate for DNN verification)

### 2024-01-27 Paper/introduction.tex

\item Our third primary contribution addresses a practical concern by 
introducing \CMP (Algorithm \ref{algo1}), a pragmatic solution that builds on our theoretical insights. This method prioritizes ReLU nodes based on the weights of their associated compensating paths and differentiates their treatment in the MILP model, employing binary constraints for heavier nodes and linear constraints for lighter ones. Our extensive empirical evaluation (refer to Table \ref{tab:example}) reveals \CMP achieves a beneficial balance between accuracy and completeness compared to prevailing methods. It reduces the proportion of undecided inputs from 17.6\% (as seen with $\beta$-Crown \cite{crown}) to 10.4\% on the DNN architecture of $8 \times 200$. Moreover, it verifies over 20\% more images than $\alpha,\beta$-CROWN \cite{xu2020fast} (or PRIMA \cite{prima}) on the larger $6 \times 500$ DNN.

@@

\section{Comparison with Related Works}

We compare CMP with several (but not all) main verification tools for DNNs, to better explain our methodology and how it differs with the existing SOTA. Compared with the exact encoding of a DNN using MILP \cite{MILP}, 
we select a restricted number of ReLU nodes based on compensating strength to scale to larger DNNs while maintaining good accuracy. Further, compared to MIPplanet \cite{MIPplanet} which 
uses a different selection of ReLU nodes, we compute bounds for all neurons iteratively, reducing drastically the number of ReLU nodes represented exactly for a given neuron,
%(ReLU nodes which were considered for neurons in previous layers are not represented exactly for that neuron, while MIPplanet would need to represent all those ReLU nodes exactly), 
using many fast computations, more efficient than one large computation.

Compared with the linear relaxation of the MILP encoding \cite{MILP}, \CMP is strictly more accurate by design, but it will also be slower.

Compared with ERAN-Deeppoly \cite{deeppoly}, which compute bounds on the value in a very efficient way, we prove that the LP encoding is strictly more accurate, which as far as we know is another novel result. To be more precise, DeepPoly
abstract the weight of every node using two functions, one upper function and one lower function. While the upper function is fixed, there are 2 choices for the lower bound.
We prove in Proposition  \ref{LP} that the LP relaxation corresponds exactly to the intersection of both choices. It is thus more accurate than DeepPoly, but also not as efficient. Therefore, \CMP will also be (much) more accurate than DeepPoly, but also not as efficient.

Concerning PRIMA \cite{prima}, the idea is to keep explicitly dependencies between neurons, computing bounds layer by layer (as we do). This allows to maintain dependencies from potentially many layers beforehand. We take care of dependencies between neurons in a different way, as compensation is the reason why there are dependencies between neurons. 
Our method is more accurate for local dependencies, but will tend to lose precision for dependencies created several layers ago. Most of the dependencies are local, in the few last layers (because ReLU nodes will likely clip those that happened many layers ago), which explains why we obtain more accurate results than PRIMA.

$\alpha,\beta$ CROWN \cite{crown} (and other Branch and Bound algorithms, such as BaBSR \cite{BaB}) will run few instances of branch and bound (one per output neuron), in worst case considering all the possible ReLU configurations (although the branch and bound algorithm avoids most of the possibilities). On simple networks, such as those trained robust, branch and bound is  particularly efficient as it consider the important branches for the actual property to verify, while \CMP is less efficient as it has to consider each node one by one from the start, including nodes whose bounds are not important. However, branch and bound faces a complexity wall when the network is hard to verify, such as the DNNs trained naturally, as there are too many branches to consider: for instance, Branch and Bound verifies only $3.5\%$ more images ($44.5\%$ from $41\%$) by enlarging the runtime by $50$ times on the larger $6 \times 5000$ DNN, and cannot close the gap with \CMP (65\% of images verified, Table \ref{tab:example}).

On such complex DNNs, $\alpha,\beta$ CROWN (and PRIMA) resorts to a "refined" path, where the bounds {\em on the first few layers} are refined \cite{MILP2} using an exact MILP encoding. In \CMP, we do not use an exact encoding but a partial one with the most important ReLU nodes obtained by considering the compensation strength. As it is more efficient, we compute bounds for neurons in all the layers. This would be infeasible without the selection based on the compensation. The refined version of $\alpha,\beta$ CROWN is particularly accurate on small DNNs. As the depth grows, the more work is left to BaB and \CMP is more accurate (Table \ref{tab:example}).
%lowering the gap of verified images to the upper bound \cite{attack} down from 
%$20\%$ to $16.2\%$ and from $17.6\%$ to $10.9\%$ (depth 8, Table \ref{tab:example}).

Finally, algorithms abstracting the network (e.g. Reluplex / Marabou \cite{Reluplex,katz2019marabou}) are very different from algorithms abstracting the values (PRIMA, ($\alpha$)($\beta$)-CROWN)\cite{prima,crown}, \CMP $\ldots$ These algorithms have been developed to be {\em complete}, so they are slower but also more accurate than what we propose.

>>>>

\item Our third primary contribution addresses a practical concern by 
introducing {\CMP} (Algorithm \ref{algo1}), a pragmatic solution that builds on our theoretical insights. This method prioritizes ReLU nodes based on the weights of their associated compensating paths and differentiates their treatment in the MILP model, employing binary constraints for heavier nodes and linear constraints for lighter ones. Our extensive empirical evaluation (refer to Table \ref{tab:example}) reveals {\CMP} achieves a beneficial balance between accuracy and completeness compared to prevailing methods. It reduces the proportion of undecided inputs from 17.6\% (as seen with $\beta$-Crown \cite{crown}) to 10.4\% on the DNN architecture of $8 \times 200$. Moreover, it verifies over 20\% more images than $\alpha,\beta$-CROWN \cite{xu2020fast} (or PRIMA \cite{prima}) on the larger $6 \times 500$ DNN.

@@

### 2024-01-27 Paper/valabstraction.tex

The idea to perform value abstraction is to compute upper and lower bounds on the values of (some) neurons in the DNN when the inputs are in a given neighbourhood, in order to conclude on the robustness without needing to compute the values exactly for all inputs (in the neighbourhood).

First, remark that weighted sums is a linear function, so it is easy to represent it explicitly. The only problematic part to represent accurately is the ReLU function. While it is fairly simple piecewise linear function with 2 modes ($x<0$ and $x \geq 0$), it is not linear. Considering its iterated impact along the different layers of a ReLU DNN is much more complex than it seems. First, let us remark that it is easy to represent exactly $\ReLU(x)$ in case $x$ is {\em stable}, that is, either it is known to be entirely positive, order
it is known to be entirely negative, as in both case there is only one known linear mode involved. Hence, the sole problem remains how to handle {\em unstable} neurons.

Consider the simpler abstraction, called the {\em Box abstraction} \cite{deeppoly}: inductively compute the bound on each neuron of the next layer independently, by considering the weighted sum over the bounds of the previous layer, then clip the lower bound at 
$\max(0,$ lower bound$)$ to represent the ReLU function, and so on.
For all $i$, let $x_i=\val_{\vx}(n_i)$, with $\vx=(x_1,x_2)$.
For instance, on the DNN example of Fig \ref{fig1}, $x_1,x_2 \in [-1,1]$, then $x_3,x_4 \in [-2,2]$, using ReLU $\hat{x}_3,\hat{x}_4 \in [0,2]$, and $x_5 \in [0,6]$ while $x_6 \in [0,2]$. 
While the bounds for $n_1, \ldots, n_4$ are exact (for all $\alpha$ 
in the range, there is an input $\vy$ such that $\val_{\vy}(n_i)=\alpha$), 
it is no more the case from the next layer (starting in $n_5,n_6$) because of potential 
dependencies between earlier neurons: $n_3$ reaches value $2$ only when $x_1=x_2=2$,
in which case $n_4$ reaches value $Val_{(2,2)}(n_4)=0$. Hence, it is impossible that $x_5=\Val_{\vx}(n_5)$ reaches $6$, as it would require both $n_3$ and $n_4$ to have value $2$.

A still extremely fast algorithm that corrects some of this issue is {\em DeepPoly} \cite{deeppoly}, also rediscovered as the {\em CROWN} algorithm \cite{crown}. Instead of 
hard bounds, for each neuron $n$ of layer $k$, two affine functions (with input at the previous layer $k-1$) are kept as lower bound and upper bound of the value of the node.
Denote $f_i \leq x_i \leq g_i$ with e.g.  
$f_{3}(x_3)=f_4(x_4)=0$ and 
$g_3(x_3) = \frac{x_3+2}{2}$,
$g_4(x_4) = \frac{x_4+2}{2}$. 
With that, we get 
$x_5 \leq g_3(x_3) + 2 g_4(x_4) = \frac{x_3 + 2x_4 + 6}{2} = \frac{3x_1 - x_2 + 6}{2}\leq 5$.

For $[\alpha,\beta]$ bounds on $x_i$, there is one optimal linear function for the upper bound $g_i(x_i)= \beta \frac{x_i-\alpha}{\beta-\alpha}$ for ReLU nodes.
There are two choices for the lower bound: $f^1_i(x_i) = 0$ or $f^2_i(x_i)=x_i$.
In the DeepPoly algorithm, the choice is made by considering $\alpha < 0 < \beta$ (for unstable neurons). If $|\alpha|\geq |\beta|$ then $f_i=f^1_i$ is used, while if $|\beta|>|\alpha|$ then $f_i=f^2_i$ is used. Denote {\em $\overline{\mbox{DeepPoly}}$} the DeepPoly algorithm such that $f^1_i$ is always used, and $f^2_i$ is never used. 
Unlike DeepPoly, {\em $\overline{\mbox{DeepPoly}}$} subsumes the {\em Box abstraction}. Indeed, taking bounds $[-0.2,5]$ on $x_i$, it gives $\ReLU(x_i) \in [0,5]$, 
while the normal DeepPoly algorithm would only allow to conclude that $\ReLU(x_i) \in [-0.2,5]$.

@@

From \cite{MILP}, for all $x \in [\alpha,\beta] \setminus 0$, there is a unique solution $(a,\hat{x})$ satisfying the constraints, and we have $\hat{x}=\ReLU(x)$ (and $a$ is 0 if $x < 0$ and 1 if $x>0$, and can be either for $x=0$). Such an encoding can be used for every (unstable) ReLU node, and optimizing its value allows to certify a given input. 
For Networks with hundreds of nodes or more, the associated MILP formula will have have hundreds of integer variables, and it will usually not be solvable efficiently.

MILP instances can be linearly relaxed into LP over-abstraction, where variables which were integers in $\{0,1\}$ (binary) are relaxed to be real in interval $[0,1]$, using otherwise the same encoding. Solving LP instances being Polynomial time, optimizing it is much more efficient, but at the price of a lack of accuracy (e.g. with not as tight bounds). This is called the {\em LP abstraction}.

As far as we know, no formal result states the relationship between the LP approximation and DeepPoly.
We prove now that the LP abstraction is strictly more accurate than the DeepPoly abstraction, as it actually corresponds to considering in DeepPoly both linear functions lower bounding the ReLU function, $ReLU(x) \geq 0$ and $ReLU(x) \geq x$, while DeepPoly considers exactly one.

\begin{proposition}
	\label{LP}
	Given $x \in [\alpha,\beta]$ with $\alpha < 0 < \beta$, the following two systems of constraints 
	1) for LP and 2) for DeepPoly with both lower bounds are equivalent:
	\vspace{-0.3cm}
	\begin{align*}
		& 1) \quad \hat{x} \geq x \quad \wedge \quad \hat{x} \geq 0, \quad \wedge \quad \hat{x} \leq \beta \cdot a \quad \wedge \quad \hat{x} \leq x-\alpha \cdot (1-a), \, a \in [0,1] \\
		\text{and} \quad  & 2)  \quad \hat{x} \geq x \quad \wedge \quad \hat{x} \geq 0 \quad \wedge \quad \hat{x} \leq \beta \frac{x-\alpha}{\beta-\alpha}
	\end{align*} 
\end{proposition}

\begin{proof}
	Obviously, the 2 lower bound constraints are the same in 1) and 2).

	It remains to prove that the 2 upper bound constraints of 1) LP corresponds to the 2) DeepPoly linear upper bound. For that, we study the upper bound function in the linear variable $a \in  [0,1]$, with $x \in [\alpha,\beta]$ fixed. We have $\hat{x}$ in 1) is upper bounded by $max_{a \in [0,1]} (min(\beta \cdot a, x - \alpha (1-a)))$, and this bound can be reached. 
	The function $min(\beta \cdot a, x - \alpha (1-a))$ reaches its maximum for 
	$\beta \cdot a = x - \alpha (1-a)$, that is 
	$(\beta - \alpha) a = x - \alpha$, i.e. 
	$ a = \frac{x - \alpha}{\beta-\alpha}$.
	It thus gives an upper bound at 
	$\beta \cdot a = \beta \frac{x - \alpha}{\beta-\alpha}$, which is exactly the upper-bound function used in DeepPoly.
	\hfill $\square$
\end{proof}

>>>>

The concept of value abstraction involves calculating upper and lower bounds for the values of certain neurons in a Deep Neural Network (DNN) when inputs fall within a specified range. This approach aims to assess the network's robustness without precisely computing the values for every input within that range.

Firstly, it's important to note that weighted sums represent a linear function, which can be explicitly expressed with relative ease. However, the ReLU (Rectified Linear Unit) function presents a challenge in terms of accurate representation. Although ReLU is a relatively straightforward piecewise linear function with two modes (one for $x<0$ and another for $x \geq 0$), it is not linear. The complexity arises when considering the compounded effects of the ReLU function across the various layers of a ReLU DNN. It's worth noting that representing $\ReLU(x)$ precisely is feasible when $x$ is "stable," meaning it's consistently positive or consistently negative, as there's only one linear mode involved in each scenario. Consequently, the primary challenge lies in addressing "unstable" neurons, where the linearity of the function does not hold consistently.

Consider the simpler abstraction, termed ``Box abstraction" \cite{deeppoly}: it inductively computes the bounds for each neuron in the subsequent layer independently. This is achieved by considering the weighted sum of the bounds from the previous layer, followed by clipping the lower bound at $\max(0,$ lower bound$)$ to represent the ReLU function, and so forth. For all $i$, define $x_i=\val_{\vx}(n_i)$, where $\vx=(x_1,x_2)$.

Taking the DNN example from Fig \ref{fig1}, assume $x_1,x_2 \in [-1,1]$. This implies that $x_3,x_4 \in [-2,2]$. After applying the ReLU function, $\hat{x}3,\hat{x}4$ are constrained to $[0,2]$, leading to $x_5 \in [0,6]$ and $x_6 \in [0,2]$. The bounds for $n_1, \ldots, n_4$ are exact, meaning for every $\alpha$ within the range, an input $\vy$ can be found such that $\val{\vy}(n_i)=\alpha$. However, this precision is lost from the next layer (beginning with $n_5, n_6$) due to potential dependencies among preceding neurons. For example, $n_3$ only attains the value $2$ when both $x_1$ and $x_2$ are $2$. In this case, $n_4$ would reach a value of $Val{(2,2)}(n_4)=0$. Consequently, it's implausible for $x_5=\Val_{\vx}(n_5)$ to reach $6$, as it would necessitate both $n_3$ and $n_4$ assuming the value $2$.

An extremely efficient algorithm that addresses some issues is "DeepPoly" \cite{deeppoly}, also independently discovered as the "CROWN" algorithm \cite{crown}. Instead of strict bounds, for each neuron $n$ in layer $k$, DeepPoly maintains two affine functions representing the lower and upper bounds of the neuron's value, based on inputs from the previous layer $k-1$. For instance, denote $f_i \leq x_i \leq g_i$ with, for example, $f_{3}(x_3)=f_4(x_4)=0$ and $g_3(x_3) = \frac{x_3+2}{2}$, $g_4(x_4) = \frac{x_4+2}{2}$. This leads to $x_5 \leq g_3(x_3) + 2 g_4(x_4) = \frac{x_3 + 2x_4 + 6}{2} = \frac{3x_1 - x_2 + 6}{2}\leq 5$.

For bounds $[\alpha,\beta]$ on $x_i$, the optimal linear function for the upper bound is $g_i(x_i)= \beta \frac{x_i-\alpha}{\beta-\alpha}$ for ReLU nodes. There are two options for the lower bound: $f^1_i(x_i) = 0$ or $f^2_i(x_i)=x_i$. DeepPoly selects between these based on the values of $\alpha$ and $\beta$ (for unstable neurons): if $|\alpha|\geq |\beta|$, then $f_i=f^1_i$ is chosen, otherwise, for $|\beta|>|\alpha|$, $f_i=f^2_i$ is selected. The variation, {\em $\overline{\mbox{DeepPoly}}$}, consistently chooses $f^1_i$, and never $f^2_i$. Contrary to DeepPoly, {\em $\overline{\mbox{DeepPoly}}$} encompasses the ``Box abstraction." For instance, with bounds $[-0.2,5]$ on $x_i$, it deduces $\ReLU(x_i) \in [0,5]$, whereas  DeepPoly without box abstraction algorithm would conclude $\ReLU(x_i) \in [-0.2,5]$.

@@

It can be shown that for all $x \in [\alpha,\beta] \setminus 0$, there exists a unique solution $(a,\hat{x})$ that meets these constraints, with $\hat{x}=\ReLU(x)$. Here, $a$ is 0 if $x < 0$, 1 if $x>0$, and can be either if $x=0$~\cite{MILP}. This encoding approach can be applied to every (unstable) ReLU node, and optimizing its value can help in certifying a given input. However, for networks with hundreds of nodes or more, the resulting MILP formulation will contain numerous integer variables and generally cannot be solved efficiently.

MILP instances can be linearly relaxed into LP over-abstraction, where variables originally restricted to integers in ${0,1}$ (binary) are relaxed to real numbers in the interval $[0,1]$, while maintaining the same encoding. As solving LP instances is polynomial time, this optimization is significantly more efficient. However, this efficiency comes at the cost of precision, often resulting in less stringent bounds. This approach is termed the {\em LP abstraction}

The following proposition shows that the LP abstraction is strictly more precise than the DeepPoly abstraction. This is attributed to the LP abstraction's simultaneous consideration of the two linear lower bounding functions for the ReLU operation, namely $ReLU(x) \geq 0$ and $ReLU(x) \geq x$. Conversely, the DeepPoly abstraction restricts itself to the application of a single one of these linear bounding functions.

\begin{proposition}
	\label{LP}
	Given $x \in [\alpha,\beta]$ with $\alpha < 0 < \beta$, the following two systems of constraints 
	1) for LP and 2) for DeepPoly with both lower bounds are equivalent:
	\begin{align}
& \hat{x} \geq x \quad \wedge \quad \hat{x} \geq 0, \quad \wedge \quad \hat{x} \leq \beta \cdot a \quad \wedge \quad \hat{x} \leq x-\alpha \cdot (1-a), \, a \in [0,1]  \label{eq:lp}\\
&\hat{x} \geq x \quad \wedge \quad \hat{x} \geq 0 \quad \wedge \quad \hat{x} \leq \beta \frac{x-\alpha}{\beta-\alpha} \label{eq:deeppoly}
	\end{align} 
\end{proposition}

\begin{proof}
We need to show that the two upper bound constraints in Eq~\ref{eq:lp} is equivalent to the upper bound in Eq~\ref{eq:deeppoly}. To this end, we focus on the upper bound function in the linear variable $a \in  [0,1]$, with $x \in [\alpha,\beta]$ fixed. We have $\hat{x}$ in Eq~\ref{eq:lp} is upper bounded by $max_{a \in [0,1]} (min(\beta \cdot a, x - \alpha (1-a)))$, and this bound can be reached. Furthermore, 
	The function $\min(\beta \cdot a, x - \alpha (1-a))$ attains its maximum when $\beta \cdot a = x - \alpha (1-a)$, leading to the equation $(\beta - \alpha) a = x - \alpha$ and consequently $a = \frac{x - \alpha}{\beta-\alpha}$. This results in an upper bound of $\beta \cdot a = \beta \frac{x - \alpha}{\beta-\alpha}$

\end{proof}
